Product plots

### 2011-03-13 prodplots-infovis.tex
Improving paper further

\begin{table*}[htb]
  \begin{center}
  \begin{tabular}{llp{9cm}}
    \toprule
    Variable & Description & Values \\
    \midrule
    {\sf age} & age in years & 18--89 \\
    {\sf degree} & highest education & lt high school, high school, junior college, bachelor, graduate \\
    {\sf finrela} & relative financial status & far above average, above average, average, below average, far below average \\
    {\sf happy} & happiness & \key{very-happy} very happy, \key{pretty-happy} pretty happy, \key{not-too-happy} not too happy \\
    {\sf health} & health & excellent, good, fair, poor \\
    {\sf sex} & sex & \key{female} female, \key{male} male\\
    {\sf wtsall} & probability weight & 0.43--6.42 \\
    {\sf year} & year of survey & 1972--2006 \\
    \bottomrule
  \end{tabular}
  \end{center}
  \caption{Description of sample data. Common plot colours are shown next to respective levels.}
  \label{tbl:happy}
\end{table*}

@@

It is useful to consider the constraints on our visualisation to determine the best solution to the problem \citep{keim:2002}. All area plots correspond to tilings of the plane. We could consider partitions of higher-d spaces (e.g. 3d or 4d), but given that we have to project these down to 2d for viewing on paper or screen, there is little disadvantage to working directly in 2d.

To derive our graphical primitives we start with three constraints:

@@

  \item Area must be {\bf proportional} to value. Because the total area for a graphic is usually constrained, an area plot partitions the total, displaying proportions rather than counts.

@@

@@

We can create two primitives by combining these one primitives, as shown in Section~\ref{sec:combination}, but there is one 2d primitive that does not arise in this way.

@@

A special case of the fluct is the equal bin size plot \citep{hofmann:2000} which occurs when the two variables are jointly uniformly distributed, usually as a result of the conditioning described in the following section. The equal bin size plot is particularly useful as a way visualise missing combinations.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\linewidth]{part-fluct}%
    \includegraphics[width=0.5\linewidth]{part-fluct-cond}
  \caption{(Left) A fluct showing distribution of age (in decades) and marital status. (Right) Equal bin-size plot showing health status and survey year. Health status was not recorded for three years.}
  \label{fig:fluct}
\end{figure}

\section{Factorising area plots}
\label{sec:combination}

To construct plots of higher-dimensional data sets, we need a way to decompose them into 1d and 2d components. Some statistical vocabulary is useful. For example, let $f(x, y, z)$ be a 3d probability mass function. A probability mass function is just a way of describing the table of counts or proportions as a function: the three parameters are three indices, saying where to look up the number. We can write this pmf as product of 1d and 2d marginals and conditionals.

  \item Joint distribution:

  \item Marginal distribution is just a name for a lower dimensional projection of a distribution - in a different context, these could be joint distributions. For example, in the three d case, there are three 2d marginal distributions: $f(x, y)$, $f(x, z)$ and $f(y, z)$, and three 1d marginal distributions, $f(x)$, $f(y)$ and $f(z)$.

  \item A conditional distribution is a slice of a higher dimension distribution, fixing one or more variables at a particular value.

\begin{figure*}[ht]
\centering
\begin{tabular}{rrr}
  \toprule
  & male & female \\ 
  \midrule
  not too happy & 0.05 & 0.07 \\ 
  pretty happy & 0.25 & 0.31 \\ 
  very happy & 0.14 & 0.18 \\ 
  \bottomrule
\end{tabular}\hspace{2em}%
\begin{tabular}{rrrr}
  \toprule
  & male & female & {\bf margin} \\ 
  \midrule
  not too happy & 0.43 & 0.57 & \hspace{2em} 0.12 \\ 
  pretty happy & 0.45 & 0.55 & 0.56 \\ 
  very happy & 0.43 & 0.57 & 0.32 \\ 
  \bottomrule
\end{tabular}\hspace{1em}%
\begin{tabular}{rrr}
  \toprule
  & male & female \\ 
  \midrule
  not too happy & 0.12 & 0.12 \\ 
  pretty happy & 0.57 & 0.55 \\ 
  very happy & 0.31 & 0.32 \\[1em] 
  {\bf margin} & 0.44 & 0.56 \\ 
  \bottomrule
\end{tabular}
  \caption{The distribution of happiness and sex, displayed in three equivalent ways.  (Left) Joint distribution. Overall table sums to one. (Middle) Conditional distribution of sex given happiness and marginal distribution of happiness. (Right) Conditional distribution of happiness given sex and marginal distribution of sex.}
\label{fig:fluct}

\end{figure*}

This is easy to see in two dimensions.  Figure XXX shows three representations of the

An important property of the definition of conditional probability is that it means you can write a joint distribution as a product of conditional and marginals:

\begin{itemize}
  \item $f(x, y, z) = f(z) f(x, y | z)$
  \item $f(x, y, z) = f(y, z) f(x | y, z) $
  \item $f(x, y, z) = f(z) f(y | z) f(x | y, z)$
\end{itemize}

This means that if we can translate this multiplication into some kind of graphical product, we can display any nD data set using low-D primitives.

\noindent Since each component of the product is only 1 or 2d, we can use one of the primitives described above. For example, Figure \ref{fig:recursive} shows two ways to display the joint distribution of sex and happiness as broken down into the marginal distribution of happiness, and the condition distribution of sex, given happiness.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.33\linewidth]{part-comb-1}%
    \includegraphics[width=0.33\linewidth]{part-comb-2}%
    \includegraphics[width=0.33\linewidth]{part-comb-3}
  \caption{Conditional on marital status, are men or women happier?  This figure shows the construction of $\sim happy + sex + marital$ from (left to right) vspines by marital status, vspines by marital status and sex, and hspines by marital status, sex and happiness.  For all levels of marital status, men are slightly less happy.}
  \label{fig:recursive}
\end{figure}

% Note that while $f(x, y, z)$ and $f(z, y, x)$ represent the same pmf, the factorisation will be different because you will get different marginal tables.

Writing the full pmf is tedious, so we define an algebra based on the modelling specification of \citet{wilkinson:1973}:

\begin{itemize}
  \item \verb|~ a| represents $f(a)$,
  \item \verb|~ a + b| represents $f(a, b)$
  \item \verb!~ a | b! represents $f(a | b)$
  \item \verb!~ . | a! is a special case, $f(a) = \frac{1}{\#A}$.
\end{itemize}

Figure~\ref{fig:fact-simple} uses the distribution of sex and happiness to illustrate these ideas. From left to right, we have plots of \verb|~ happy|, \verb!~ happy | sex! and the product \verb|~ happy + sex|. Observe that the heights and widths of the rectangles are multiplied in the same way as the components of the pmf.

@@

@@

Different combinations of partitions reveal different features of the data. Take for example, the distribution of age and marital status, as shown in Figure~\ref{fig:fluct}. Instead of visualisation the joint distribution with a fluct, we could focus on the conditional distribution of marital status, given age, or vice versa. Figure~\ref{fig:marital} shows two ways to do this. The left plot shows $\sim marital + age$ with a vspine nested in a hspine, and the right plot $\sim age + martial$ with a hbar nested in a vspine. These displays show the same data but are arranged in such a way to make different comparisons easier.

@@

Conditioning is also an important tool by itself, because it allows one to remove relationships that are well-known or uninteresting. Figure~\ref{fig:part-cond} uses a fluct and a vspine to explore the relationship between happiness, health and financial status. The left plot displays raw proportions, showing that most people are in good health and average financial standing. However, it is difficult to see how happiness varies within these conditions because the perceptual task is comparing areas. Conditioning on financial status and health produces the plot on the right (an equal bin size plot) and makes it easier to see the conditional distribution of happiness, because the perceptual task is simpler: comparing length on a common non-aligned scales.  Here we can see people tend to be happier the healthier and richer they are.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\linewidth]{part-fluctuation}%
    \includegraphics[width=0.5\linewidth]{part-equal-area}
  \caption{(Left) $\sim happy + health + finrela$, partitioned with a vspine and fluct.  (Right) $\sim happy \,|\, health + finrela$. We can no longer see the joint distribution of health and financial status, but it is much easier to see the conditional distribution of happiness. Healthier and richer people are happier: maybe money does buy happiness? (\key{not-too-happy} Not too happy, \key{pretty-happy} pretty happy, \key{very-happy} very happy)}
  \label{fig:part-cond}
\end{figure}

@@

You may have noticed that the previous plots resemble area plots that have previously be named. The one-d primitives already have their own names:

\begin{itemize}
  \item {\bf Bar chart} (1d). 1 hbar.
  \item {\bf Column chart} (1d). 1 vbar.
  \item {\bf Spineplot} (1d). 1 spine \citep{hummel:1996}.
  \item {\bf Fluctuation} diagram (2d): 1 fluct.
\end{itemize}

@@

\begin{itemize}
  \item {\bf Stacked} bar chart (2d). 1 hbar and 1 vspine.

  \item {\bf Nested} bar chart (2d).  2 hbars. \citep{peltier:2009}

  \item {\bf Equal bin size} plot (3d): fluct and vspine, conditioned on the first two variables \citep{hofmann:2000}.

  \item {\bf Mosaic} plot (nd).  Alternating hspines and vspines.  \citep{hartigan:1981,friendly:1994,hofmann:2003}

  \item {\bf Double-decker} plot (nd).  $n-1$ hspines and 1 vspine. \citep{hofmann:2001}

  \item {\bf Treemap} (nd): n spines. \citep{shneiderman:1992}

  \item {\bf Squarified treemap} (nd): n tiles. \citep{bruls:1999}

  \item {\bf Generalised treemaps} (nd): any plot ending with a tile \citep{vliegen:2006}.

@@

On top of this basic framework, it is useful to consider a few variations and extensions.  Three are based on variations of the underlying data (missing data, weighted data, and continuous data), and three are based on relaxing the display constraints (area not proportional to weight, non-disjoint partitions, non-rectangular partitions).

\subsection{Missing values and zeros}
\label{sub:missing_values}

Crossed vs. nested

@@

We have assumed that the probabilities represent counts, but without loss of generality, we can use any set of non-negative, additive weights. For example, in the happy dataset, the {\sf wtssall} variable gives analytic survey weights. Figure~\ref{fig:weighting} shows the difference between the weighted and unweighted distributions of age and sex. The distribution is subtly different. \citep{unwin:2007} describes the applications of weighted plots in more detail.

To express that we are displaying weights, rather than counts, we extend the formula syntax by putting the weighting variable on the left-hand side: $wtssall \sim sex + age$ vs. $ \sim sex + age$.

@@

These techniques can also be used with continuous data, if the data is binned to create a categorical variable. There are many different possible choices of binning algorithms, but we will focus on two:

\begin{itemize}
  \item Intervals of equal width.
  \item Intervals containing equal numbers of points.
\end{itemize}

This gives rise to the histogram and spinogram \citep{hummel:1996}, continuous analogues of the bar chart and spine chart. A long standing tradition is that no gaps are displayed between adjacent rectangles when used for continuous data. Two examples are shown in Figure \ref{fig:cont-examples}.

\begin{figure}[htbp]
  \centering
  \caption{Area plots of originally continuous data}
  \label{fig:cont-examples}
\end{figure}

Example exploring joint and conditional distributions of age and year.

If an equal bin sized plot is used, and value is mapped to colour, then we get  dimensional stacking \citep{leblanc:1990}

% An alternative approach is use density estimates (or conditional density estimates) and plot with continuous area plots.  As in: Hofmann, H., Theus, M. (2005), Interactive graphics for visualizing conditional distributions, Unpublished Manuscript.

@@

It can be useful to violate the constraint that area should be proportional to weight to distinguish between zeros and very small values. A zero weight should have zero area, but giving it positive area can be useful, so you can actually see it! In general, it's useful to constrain all areas to be above a certain minimal perceptible size.  Areas which are constrained in such a way should have a visual flag (such as a difference colour) so ensure that the reader knows that the relationship between area and value has been violated.

On the other end of the spectrum, it can be useful to constrain put a cap on the largest possible values \citetext{Antony Unwin, priv. comm.}. By controlling this cap it is possible to zoom in on the smallest values. - called censored zooming by Antony and Martin.

Approach is also taken by \citep{keim:2002}, where a minimum bar width of 10 pixels is used.

Other non-linear transformations may also be useful. For example, we could display the square root of the weights, to stabilise the variance of the areas. Tukey applied this technique to histograms to create rootograms.

If values span multiple orders of magnitude it may be useful to map log value to area.

@@

The cascaded treemaps of \citet{lu:2008} is an idea that illustrates how the violation of containment can be productive.  In the cascaded treemap, each level is slightly offset from the one above to create a pseudo-3d perspective.
This makes it easier to see all the levels of the hierarchy, not just the lowest level.  Figure~\ref{fig:cascade} shows an example of how cascading can help illuminate the structure of a complex mosaic plot.

@@

Many other circular displays turn out to be special cases of product plots drawn in polar coordinates. Generally, the y axis (mapped to radius) must be square-root transformed to ensure that that counts stay proportional to areas. Stasko's radial displays \citep{stasko:2000} and fan-lenses \citep{lou:2007} deliberately do not do this in order to emphasise the outer levels.

@@

\begin{itemize}
  \item The wind rose \citep{lalanne:1843}, fourfold displays \citep{friendly:1995}, and sector graphic use the hbar. Nightingales's coxcomb \citep{nightingale:1857} is very similar, but the slices overlap and so violate the constraint of disjoint area.

  \item The bullseye chart uses a hspine.

  \item The doughnut plot, as popularised by Excel, uses a vspine and a hspine.

  \item The circular bar chart or racetrack plot, uses a vbar.

  \item Infoslices \citep{andrews:1998} use a nested vbar. But they only use half of the polar plane, and are specialised for highly nested data.

@@

\section{Aesthetics}
\label{sec:aesthetics}

\subsection{Labelling} 
\label{sub:labelling}

\subsection{Offsets}
\label{sub:offsets}

\section{Interaction}
\label{sec:interaction}

@@

\noindent There are 24 possible ways of ordering the variables in the pdf, and each 1d pmf can be displayed in five different ways. For a 4d problem this gives $5 * 24 * (5^3 + 5^4 + 5^2 + 1 + 5^2) = 96120$ plots!  In future work, we hope to connect product plots to statistical models to help guide the choice of appropriate graphic.

>>>>

\begin{table*}[htb]
  \begin{center}
  \begin{tabular}{llp{9cm}}
    \toprule
    Variable & Description & Values \\
    \midrule
    {\sf age} & age in years & 18--89 \\
    {\sf degree} & highest education & lt high school, high school, junior college, bachelor, graduate \\
    {\sf finrela} & relative financial status & far above, above average, average, below average, far below \\
    {\sf happy} & happiness & \key{very-happy} very happy, \key{pretty-happy} pretty happy, \key{not-too-happy} not too happy \\
    {\sf health} & health & excellent, good, fair, poor \\
    {\sf sex} & sex & \key{female} female, \key{male} male\\
    {\sf wtsall} & probability weight & 0.43--6.42 \\
    {\sf year} & year of survey & 1972--2006 \\
    \bottomrule
  \end{tabular}
  \end{center}
  \caption{Description of sample data. Common plot colours are shown next to respective levels.}
  \label{tbl:happy}
\end{table*}

@@

All area plots correspond to tilings of the plane. We could consider partitions of higher-d spaces (e.g. 3d or 4d), but given that we have to project these down to 2d for viewing on paper or screen, there is little disadvantage to working directly in 2d. There are many possible ways to tile the plane. To weed these down to a manageable number, we identify three constraints \citep{keim:2002} important for visualising counts:

@@

  \item Area must be {\bf proportional} to count. Because the total area for a graphic is usually constrained, and area plot partitions the total, area plots display proportions, not counts. From here on, without loss of generality, we will refer to visualisation of proportions, rather than counts.

@@

\begin{itemize}
   \setlength{\itemsep}{0em}

@@

We can create 2d primitives by combining these 1d primitives, as shown in Section~\ref{sec:combination}, but there is one 2d primitive that does not arise in this way.

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.9 \linewidth]{part-fluct}
    \includegraphics[width=0.9 \linewidth]{part-fluct-cond}
  \caption{(Left) A fluctuation diagram showing distribution of age (in decades) and marital status. (Right) Equal bin-size plot showing health status and survey year. Health status was not recorded for three years.}
  \label{fig:fluct}
\end{figure}

A special case of the fluct is the equal bin size plot \citep{hofmann:2000} which occurs when the two variables are jointly uniformly distributed, usually as a result of the conditioning described in the following section. The equal bin size plot is particularly useful as a way of visualising missing combinations. Figure~\ref{fig:fluct} illustrates these two types of graphics.

\section{Probability and plot products}
\label{sec:combination}

To construct plots of higher-dimensional data sets, we need a way to decompose them into 1d and 2d components. Some statistical vocabulary is useful. One way of describing the input data is as a {\bf probability mass function}, or {\sc pmf}. A {\sc pmf} is a function with n inputs, each indexing one dimension with an integer, and gives the probability of each combination of inputs. A {\sc pmf} has two restrictions: every value must be greater than or equal to zero, and all value must sum to one.

\subsection{Products of probabilities}

Figure~\ref{fig:2d-table} shows three ways to represent the the 2d table of proportions of sex and happiness. The top table displays the {\bf joint distribution}, and allows us to answer questions of the form ``what proportion of all people are male and very happy?'' The middle two tables displays two {\bf conditional distributions}: the distribution of sex given happiness, and happiness given sex. These correspond to restriction the row or column sums to equal one, and support questions such as ``what proportion of pretty happy people are female?'' or ``what proportion of males are not too happy?''. Finally, the last table display the two {\bf marginal distributions} of sex and happiness. These allow us to answer questions like ``what proportion of respondents were male'' or ``what proportion of respondents were very happy?''.

\begin{figure}[ht]
  \begin{tabular}{rrr}
  \toprule
  & male & female \\ 
  \midrule
  not too happy & 0.05 & 0.07 \\ 
  pretty happy & 0.25 & 0.31 \\ 
  very happy & 0.14 & 0.18 \\ 
  \bottomrule
  \end{tabular}
  \\[1em]

  \begin{tabular}{rrr}
  \toprule
  & male & female \\
  \midrule
  not too happy & 0.43 & 0.57 \\
  pretty happy & 0.45 & 0.55 \\ 
  very happy & 0.43 & 0.57 \\
  \bottomrule
  \end{tabular}\hspace{1em}%
  \begin{tabular}{rrr}
  \toprule
  & male & female \\ 
  \midrule
  not too happy & 0.12 & 0.12 \\ 
  pretty happy & 0.57 & 0.55 \\ 
  very happy & 0.31 & 0.32 \\
  \bottomrule
  \end{tabular}
  \\[1em]

  \begin{tabular}{rrrr}
  \toprule
  & male & female & \\ 
  \midrule
  not too happy & & & 0.12 \\ 
  pretty happy & & & 0.56 \\ 
  very happy & & & 0.30 \\
  & 0.44 & 0.56 \\ 
  \bottomrule
  \end{tabular}

  \caption{The distribution of happiness and sex, displayed in three equivalent ways.  (Left) Joint distribution. Overall table sums to one. (Middle) Conditional distribution of sex given happiness and marginal distribution of happiness. (Right) Conditional distribution of happiness given sex and marginal distribution of sex.}
  \label{fig:2d-table}
\end{figure}

Formally, the conditional distribution is written as $f(x | y)$ and is equal to $f(x, y) / f(y)$. This definition illustrates that given a conditional distribution and a marginal distribution, we can always find the joint distribution: $f(x, y) = f(x | y) f(y)$ \footnotemark. This definition extends in a straightforward way to higher dimensions. For example, a 3d pmf can be written as the product of 2d and 1d conditional and marginal distributions as follows:

\footnotetext{Because the parameters of a pmf are identifying, statisticians often fail to distinguish the different functions.  To be precise, the above statement should be written: $f_{X, Y}(x, y) = f_{(X|Y=y)}(x | y) f_Y(y)$}

\begin{itemize}
  \setlength{\itemsep}{0em}
  \item $f(x, y, z) = f(z) f(x, y | z)$
  \item $f(x, y, z) = f(y, z) f(x | y, z) $
  \item $f(x, y, z) = f(z) f(y | z) f(x | y, z)$
\end{itemize}

This means that we can turn any high-dimensional {\sc pmf} into a product of low-dimensional conditional and marginal {\sc pmf}s products. It remains to find an graphical analog of this multiplication.

\subsection{Area is a product of height and width}

We connect probability products to our rectangular primitives by noting that areas are also products, products of height and width. It's easiest to show this with a picture: Figure~\ref{fig:fact-simple} shows how our simple one primitives combine to get two familiar plots: the mosaic plot and the stacked bar chart. From left to right, we have plots of $f(happy)$, $f(happy | sex)$ and the product $f(happy, sex)$. The heights and widths of the rectangles are multiplied in the same way as the components of the {\sc pmf}.

@@

Figure \ref{fig:recursive} shows the steps along by which a 3d dataset is displayed.  We first display $f(martial status)$, then $f(marital sex, sex) = f(sex | marital status) f(marital status)$, and finally $f(marital sex, sex, happiness) = f(happiness | sex, marital status) f(sex | marital status) f(marital status)$.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.33\linewidth]{part-comb-1}%
    \includegraphics[width=0.33\linewidth]{part-comb-2}%
    \includegraphics[width=0.33\linewidth]{part-comb-3}
  \caption{Conditional on marital status, are men or women happier?  This figure shows the construction of $f(happy, sex, marital)$ from (left to right) vspines by marital status, vspines by marital status and sex, and hspines by marital status, sex and happiness.  For all levels of marital status, men are slightly less happy.}
  \label{fig:recursive}
\end{figure}

% Note that while $f(x, y, z)$ and $f(z, y, x)$ represent the same pmf, the factorisation will be different because you will get different marginal tables.
% 
% Writing the full pmf is tedious, so we define an algebra based on the Wilkinson's modelling specification \citep{wilkinson:1973}:
% 
% \begin{itemize}
%   \setlength{\itemsep}{0em}
%   \item \verb|~ a| represents $f(a)$,
%   \item \verb|~ a + b| represents $f(a, b)$
%   \item \verb!~ a | b! represents $f(a | b)$
%   \item \verb!~ . | a! is a special case, $f(a) = \frac{1}{\#A}$.
% \end{itemize}
%

@@

Different combinations of partitions reveal different features of the data. Take for example, the distribution of age and marital status, as shown in Figure~\ref{fig:fluct}. Instead of visualisation the joint distribution with a fluct, we could focus on the conditional distribution of marital status, given age, or vice versa. Figure~\ref{fig:marital} shows two ways to do this. The left plot shows $f(marital, age)$ with a vspine nested in a hspine, and the right plot $f(age, marital)$ with a hbar nested in a vspine. These displays show the same data but are arranged in such a way to make different comparisons easier. The second plot, with bars nested inside spines, also illustrates important feature of non-space filling tilings: the relationship between proportion and area is only constant within a level. The bars are scaled to be as high as possible, without overflowing the bounding area.

@@

Conditioning is also an important tool by itself, because it allows one to remove relationships that are well-known or uninteresting. Figure~\ref{fig:part-cond} uses a fluct and a vspine to explore the relationship between happiness, health and financial status. The left plot displays raw proportions, showing that most people are in good health and average financial standing. However, it is difficult to see how happiness varies within these conditions because the perceptual task is comparing areas. Conditioning on financial status and health produces the plot on the right (an equal bin size plot) and makes it easier to see the conditional distribution of happiness given sex and health, because the perceptual task is simpler.  Depending on the comparison we are most interested in, we can make it easier to compare across wealth for fixed health, or health for fixed wealth.

For this data, for a fixed income level, happiness always increases as health increases. The same is not true for fixed health: those with poor or fair health are actually less happy if they are far above average, compared to just above average.

\begin{figure*}[htbp]
  \centering
    \includegraphics[width=0.33\linewidth]{part-fluctuation}%
    \includegraphics[width=0.33\linewidth]{part-equal-area}%
    \includegraphics[width=0.33\linewidth]{part-equal-area-2}
  \caption{(Left) $f(happy, health, finrela) = f(happy | health, finrela) \times f(health, finrela)  $, partitioned with a vspine and fluct.  (Middle) $f(happy | health, finrela)$. We can no longer see the joint distribution of health and financial status, but it is much easier to see the conditional distribution of happiness. Healthier and richer people are happier: maybe money does buy happiness? (Right) $f(happy | health, finrela)$ partitioned with a fluct and {\bf h}spine. (\key{not-too-happy} Not too happy, \key{pretty-happy} pretty happy, \key{very-happy} very happy)}
  \label{fig:part-cond}
\end{figure*}

@@

Many existing plots fall into this framework. The primitives already have their own names:

\begin{itemize}
  \setlength{\itemsep}{0em}
  \item {\bf Bar chart} (1d). 1 hbar.
  \item {\bf Column chart} (1d). 1 vbar.
  \item {\bf Spineplot} (1d). 1 spine.
  \item {\bf Fluctuation} diagram (2d): 1 fluct.
\end{itemize}

@@

\begin{itemize}
  \setlength{\itemsep}{0em}

  \item {\bf Stacked} bar chart (2d). 1 hbar and 1 vspine.

  \item {\bf Nested} bar chart \citep{peltier:2009} (2d).  2 hbars.

  \item {\bf Equal bin size} \citep{hofmann:2000} plot (3d): fluct and vspine, conditioned on the first two variables.

  \item {\bf Mosaic} plot \citep{hartigan:1981,friendly:1994,hofmann:2003} (nd). Alternating hspines and vspines.

  \item {\bf Double-decker} plot \citep{hofmann:2001} (nd). $n-1$ hspines and 1 vspine.

  \item {\bf Treemap} \citep{shneiderman:1992} (nd): n spines.

  \item {\bf Squarified treemap} \citep{bruls:1999} (nd): n tiles.

  \item {\bf Generalised treemaps} \citep{vliegen:2006} (nd): any plot ending with a tile.

@@

On top of this basic framework, it is useful to consider a few variations and extensions. Instead of count of discrete levels, we can plot weighted data or continuous data, or we can relax the display constraints to allow displays where area is not proportional to weight, partitions are non-disjoint or non-rectangular.

@@

We have assumed that the proportions represent counts, but without loss of generality, we can use any set of non-negative, additive weights. For example, in the happy dataset, the {\sf wtssall} variable gives analytic survey weights. These are use to account for oversampling of black respondents in certain years, and to reduce the effect of non-response by other demographics. Figure~\ref{fig:weighting} shows the difference between the weighted and unweighted distributions of age and sex. The distribution is subtly different, and suggests that we don't need to worry about weights for this plot. In other datasets, weights can be useful to move from numbers of counties to numbers of people, or to areas, or to other relevant quantities. Some examples of weighted data graphics can be found in \citep{unwin:1998,unwin:2003aa,unwin:2006}.

@@

The framework can be trivially extended to work with continuous data: just first bin continuous variables. There are many different to ways to create intervals for continuous data, but two are most important \citep{denby:2009}: intervals of equal width, and intervals containing equal numbers of points. This extension allows the product plot framework to also describe {\bf histograms} and {\bf spinograms} \citep{hummel:1996}, continuous analogues of the bar and charts. A long standing tradition is that no gaps are displayed between adjacent rectangles when used for continuous data. Two examples are shown in Figure \ref{fig:cont-examples}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.5\linewidth]{cont-hbar}%
  \includegraphics[width=0.5\linewidth]{cont-hspine}

  \caption{The distribution of happyness with age. (Left) hbar + vspine: an extension of the histogram. (Right) hspine + vspine: an extesion of the spinogram.}
  \label{fig:cont-examples}
\end{figure}

With one more extension, displaying the final probability with colour, not area, this allows us to describe {\bf dimensional stacking} \citep{leblanc:1990} as a plot of conditioned, binned continuous data, with proportion shown with colour.

@@

It can be useful to violate the constraint that area should be proportional to weight to distinguish between zeros and very small values. A zero weight should have zero area, but giving it positive area can be useful that so you can actually see it! In general, it's useful to constrain all areas to be above a certain minimal perceptible size. Areas which are constrained in such a way should have a visual flag (such as a difference colour) so ensure that the reader knows that the relationship between area and value has been violated. This type of non-linear mapping has been implemented in {\sc manet} \citep{unwin:1996} and hierarchical pixel bar charts \citep{keim:2002}.

On the other end of the spectrum, it can be useful to constrain put a cap on the size of largest values to get censored zooming \citetext{Antony Unwin, priv. comm.}. Controlling this cap makes it possible to focus on the smallest values.

Other non-linear transformations may also be useful. For example, we could display the square root of the weights, to stabilise the variance of the areas. Tukey applied this technique to histograms to create rootograms \citep{wainer:1974,tukey:1977}: histograms where the y-axis has a square root scale.

@@

The cascaded treemaps of \citet{lu:2008} is an idea that illustrates how the violation of containment can be productive. In the {\bf cascaded treemap}, each level is slightly offset from the one above to create a pseudo-3d perspective. This makes it easier to see all the levels of the hierarchy, not just the lowest level. Figure~\ref{fig:cascade} shows an example of how cascading can help illuminate the structure of a complex mosaic plot.

@@

Many other circular displays turn out to be special cases of product plots drawn in polar coordinates \citep{draper:2009}. Generally, the y axis (mapped to radius) must be square-root transformed to ensure that that counts stay proportional to areas. Stasko's radial displays \citep{stasko:2000} and fan-lenses \citep{lou:2007} deliberately do not do this in order to emphasise the outer levels.

@@

\begin{itemize}
  \setlength{\itemsep}{0em}

  \item {\bf Wind rose} (aka sector graphic) \citep{lalanne:1843} and {\bf fourfold displays} \citep{friendly:1995} (2d): 1 hbar, and 1 vspine. Nightingales's coxcomb \citep{nightingale:1857} is very similar, but the slices overlap and so violate the constraint of disjoint area.

  \item {\bf Concentric pie chart} (aka bullseye chart) (1d): 1 hspine.

  \item {\bf Doughnut plot} (2d): 1 hspine, and 1 vspine.

  \item {\bf Racetrack plot} (aka circular bar chart) (1d): 1 vbar.

  \item {\bf Infoslices} \citep{andrews:1998} (nd): n vbars. But they only use half of the polar plane, and are specialised for highly nested data.

@@

% \section{Aesthetics}
% \label{sec:aesthetics}
% 
% \subsection{Labelling} 
% \label{sub:labelling}
% 
% \subsection{Offsets}
% \label{sub:offsets}
% 
% 
% \section{Interaction}
% \label{sec:interaction}

@@

\noindent There are 24 possible ways of ordering the variables in the pdf, and each 1d pmf can be displayed in five different ways. For a 4d problem this gives $5 * 24 * (5^3 + 5^4 + 5^2 + 1 + 5^2) = 96,120$ plots! In future work, we hope to connect product plots to statistical models to help guide the choice of appropriate graphic.

We are also interested in exploring the automated construction of a graphic to answer specific questions, and for given a plot, suggesting which type of questions it is most suitable to answer.
